Investment Strategy Based on Artificial Earthworm Behavior for Algotrading Application

### 2014-05-12 Article/Article20140406Granada/LatexArt/article_main.tex
przeniesienia do nowych linii: and, or....

\mainmatter  % start of an individual contribution

@@

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Antoni Wiliński
\and Aneta Bera\and Maciej Jarlaczyński\and Piotr Błaszyński}
%
\authorrunning{A. Wilinski et al}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{West Pomeranian University of Technology, Faculty of Computer Science,\\
Zolnierska 49, 71210 Szczecin, Poland\\
\mailsa\\
\url{http://wi.zut.edu.pl}}

%
% NB: a more complex sample for affiliations and the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%

\toctitle{The investment strategy based on behavior of artificial earthworm for use in algotrading}
\tocauthor{Antoni Wilinski}
\maketitle

\begin{abstract}
The paper presents an innovative concept of investment strategy derived from the general ideas of artificial intelligence. The strategy has been tested in a number of simulations on historical data as well as data outside learning periods. Tests were performed mainly in the selected currency markets, including the primary currency pair of EURUSD. Presented approach opens long and short positions alternately during candle closing events. Closing a position is based on recommendation which is deduced from changes in characteristics of time series. The strategy has parameters that vary through time and are adapted to a market changes. To present the strategy a metaphor of an innovative artificial earthworm that feeds on a time series was used. The earthworm has its own artificial intelligence which enables it to analyse ,,consumed'' candles, change its preferences about movement direction or to stay in neutral state. The goal was to implement presented strategy for trading platform to achieve the automatic trading effect.  The strategy was tested in two development environments --- MATLAB environment and trading platform MetaTrader after conversion of M-file into MQL4 file. Implemented strategy enables to achieve interesting results.
\keywords{investment strategy, machine learning, forecasting, time series, algotrading, mql4}
\end{abstract}

@@

Algotrading is a new rapidly growing form of trading on financial markets where investor (that makes decisions) is replaced with trading bot (computer program) \cite{Leshik2011,Wilinski2014}. Such machine can perform operations (opening and closing positions, placing orders to open positions that meet certain conditions) directly in an automated brokerage platform environment. It may also occur as a recommendation system that helps investor with making correct decisions.\\

Such computer applications (bots) require carefully prepared and tested software that most often needs to adapt to the changing situations on the observed markets. Modern trends, in the development of this software, include sophisticated methods of: artificial intelligence, pattern recognition, classification and machine learning \cite{Leshik2011,Wilinski,wang,sinclare}. Those methods are interpenetrating and complementary and are also expanded with information based on fundamental analysis e.g. automatically extracted from the Internet \cite{elder,Wilinski2014,Schwager1996}. Many strategies are based on relatively simple rules. Efficiency of these rules is amplified by using of machine learning \cite{Wilinski,person,tian,krutsinger,lewis,murphy}.\\

Strategy presented in this paper is an example of a solution that can be used fully automatically or with assistance of investor who makes the decision manually. Using such softwaremakes sense only if there are reasonable premises of achieving the success. These conditions can be achieved through inductive thinking \cite{provost,KleskWilinski}, involving the assumption that multiple validation of decision on data from the past substantiates the correctness of the decision in the future. If a certain set of rules in the strategy used to seek the pattern, leads to the success then these rules are a pattern that is encouraged for repeated use. If repeated use of mentioned pattern results in more successes than failures, then the basic premise to consider the use of the pattern is rational \cite{bishop2006pattern,KleskWilinski}. Unfortunately, changing market is causing the pattern to become outdated therefore it is necessary to constantly update it. Finding historically effective patterns that could be also effective in a data space not used for their determination (to identify these patterns) is a major challenge for machine learning which is applied in the paper.
The article is organized as follows: at first authors describe an authored strategy based on an entirely innovative concept of an artificial earthworm. Subsequently extensive studies of effectiveness of presented strategy in terms of prediction and the aspect of investment are carried out. At the end the conclusions resulting from the study are presented and also cited bibliographic sources are included.

\section{Strategy characteristics}
In presented strategy, a metaphorical procedure is often used to explain heuristic algorithms of artificial intelligence. The procedure is a reference to an inspiration derived from the nature.\\

Let's consider an artificial earthworm, which behaves like its natural inspirer. Let this artificial earthworm have a length of $p1$ candles (vectors representing the knowledge about events occurring within the contractual part of the time series). Let its movement come down to the successive ,,eating'' one candle and simultaneous ,,discarding'' of one candle (the last one) --- always having $p1$ candles. Thus the earthworm moves incrementally at length of time called the period of the candle. $i-th$ candle is  defined as a row of data of values  $O_i$, $H_i$, $L_i$, $C_i$ (Open, High, Low and Close). These values represent (in order): the value at the beginning of this period, the maximum value within the period, the minimum and the closing values - at the end of the candle.\\
Let's assume that candle have a color (often in a number of brokerage platforms such approach is used). Therefore candles may be red (when the opening of the candle is higher than its closing $O_i>C_i$) and green (when the situation is opposite $O_i<C_i$).\\
Let's also assume --- it is a primary strategy rule - that when the artificial earthworm has a significant dominance of red candles, a short position is opened or continued. When earthworm has the significant dominance of green candles, a long position is opened or continued. Imprecise term ,,significant'' is defined by the use of the second parameter $p2$ according to the rules:
\begin{equation}
\text{If }  Lr > p1/2 -p2 \text{ then } short;
\end{equation}
opening of short position will be suggested;
\begin{equation}
\text{If }  Lg > p1/2 -p2 \text{ then } long;
\end{equation}
opening of long position will be suggested;
\begin{equation}
\text{If }  Lg < p1/2 -p2 \text{ and } Lr< p1/2-p2 \text{ then } none;
\label{eq:eq3}
\end{equation}
the continuation of state will be suggested (position or no opened position).\\

In the rules above $Lr$ and $Lg$ are the number of candles, respectively ,,red'' and ,,green'' in the earthworm of $p1$ length, wherein $p1 = Lr + Lg$.\\
When dominance determined by another parameter $p2$ is undecided, as in (\ref{eq:eq3}), then earthworm takes the color white and the position will not be opened. When the earthworm continues to open position, investor does not incur costs. On many platforms, also in literature, symbolic marking of growing candles in white and decreasing in black is used. In this article, due to the occurrence of three values, green-red-white colors are used.\\

The earthworm is therefore supplied with artificial intellect which enables the accumulation of knowledge and decision-making:
\begin{itemize}
\item right after closing of the current candle, earthworm ,,swallows'' the candle and ,,knows'' what color is it;
\item earthworm knows how many candles it is composed with;
\item after absorbing and discarding one candle, earthworm knows how many candles inside it are red or green and how to determine its status --- red, green or white;
\item knowing its status, earthworm knows how to make a decision about opening or continuing position; the opening of a position (long or short) means the simultaneous closing of the current position (opposite) if such position is opened;
\item earthworm observes cumulative profit curve and if it  notices unacceptable drawdown, then opened position is changed to the opposite;
\item unacceptable drawdown is defined by two other parameters $p3$ and $p4$. $p3$ is the number of previous candles, which sets a reference point in time series, in which the value of the accumulated profit for $i-p3$ is checked for comparison with the current value of the accumulated profit. $p4$ is permissible decrease of cumulative profit. If it is exceeded, the $SL$ (Stop Loss --- closes the currently open position) mechanism is activated;
\item earthworm also observes current course changes during  each next candle and reacts to unprofitable changes. If the current exchange rate causes a loss on the opened positions greater than $p5$ then the $SL$ mechanism is triggered --- the position is closed and till the end of the candle earthworm remains in the state without an open position. $p5$ is therefore the fifth strategy parameter. 
\end{itemize}

These features of the earthworm's intelligence may be extended, however, for this work they were limited to the foregoing. \\

It is assumed that the time series is $S_i$ where $i = 1, 2, ...,$ and each element is a vector $S_i = (O_i, H_i, L_i, C_i)$ called candle.\\

Artificial earthworm executes the following algorithm.\\
The strategy is built on the principle of classical machine learning. Before starting to run the algorithm on data located outside the scope that investor knows (like on the current market segments --- out-of-sample), it is necessary to ,,teach''  the values of parameters. This principle is presented in Figure \ref{fig:fig2}, which shows examples of learning periods and following them testing periods. Data from testing period was not used during learning.\\
\begin{figure}[ht]
\centering
\includegraphics[width = 0.6\textwidth]{figures/rys2.png}
\caption{An example of the arrangement of learning and testing periods of strategy parameters in time series.}
\label{fig:fig2}
\end{figure}
\FloatBarrier
\vspace{-1em}
This repeatable algorithm located in both periods - learning and testing - has the following structure. In learning period the optimal value for five parameters $p1$, $p2$, ..., $p5$ is determined. In test period these designated values of parameters are used to perform the algorithm in the new environment. In each $i-th$ step (candle) of time series following activities are performed:
\begin{itemize}
\item For each $i = p1, p1 +1, p1 +2, ..., I$ and for $j = 1, ..., p1$ check whether the candle located at a distance $i-j$ backwards from the current candle is:\\
green when $O_{i-j} <C_{i-j}$, \\
red when $O_{i-j}> C_{i-j}$, \\
where $p1$ --- is the first parameter which is the length of the artificial earthworm --- the number of candles of different colors that make up the ,,body'' of the earthworm; \\
$O_{i-j}$ --- is the opening value of (i-j)-th candle;\\ 
$C_{i-j}$ --- is the closing value of (i-j)-th candle; \\
$I$ --- is the length of the test window, after which it is necessary to change the parameters values to the new, learned in the machine learning phase;\\
\item In each $i-th$ step, the loop $j = 1, ..., p1$ is performed in which a number of green $Lg$, red $Lr$ candles is calculated so the sum of $Lg$ and $Lr$ is $p1$ (to avoid the situation when $O_i = C_i$ is an arbitrary assignation to one of the candles sets e.g. green);
\item A condition for the existence of a $Rg$ recommendation to open a long position (or to continue it if it is open) is checked. This condition is arbitrarily adopted (one can of course consider different one): 
\begin{equation}
\text{If } Lg> p1/2-p2 \text{ then } Rg=1
\end{equation}

\end{itemize}
If this condition is fulfilled, then depending on the current position status $PositionState$ (where $PositionState$ is a variable which takes values from \{$long, short, none$\}) at the opening of the next candle, a position is opened or continued as follows: 
\begin{equation}
\text{If } Rg=1 \text{ and } PositionState =long \text{ then }PositionState =long;
\end{equation}
(costless continuation of opened position);
\begin{equation}
\text{If } Rg=1 \text{ and } PositionState = short \text{ then } PositionState=long;
\end{equation}
\begin{equation}
\text{If } Rg=1 \text{ and } PositionState =none \text{ then } PositionState =long;
\end{equation}

Similarly check the condition of recommendations Rr to open or continue short position:
\begin{equation}
\text{If } Lr>p1/2 - p2 \text { then Rr=1.}
\end{equation}
If this condition is fulfilled, then depending on what the current position status PositionState is, at the opening of the next candle, position is opened or continued as follows: \\
\begin{equation}
\text{If } Rr=1 \text{ and } PositionState =long \text{ then } PositionState =short;
\end{equation}
\begin{equation}
\text{If } Rr==1 \text{ and } PositionState ==short \text{ then } PositionState=short;
\end{equation}
(costless continuing of opened position)
\begin{equation}
\text{If } Rr==1 \text{ and }  PositionState ==none \text{ then }  PositionState =short;
\end{equation}

In formulas above there is a parameter $p2$. It is the second parameter of the strategy which is an arbitrarily separated part of the earthworm, which depending on size causes the direction of investment not to change. Large $p2$ causes long positions to change into short positions (or vice-versa) rarely. Small value of $p2$ leads to a higher frequency of these changes. The optimum value of $p2$ is chosen similarly like $p1$.
\begin{itemize}
\item 	When opening long position, profit after one candle (this does not mean that position was closed after a single candle) can be calculated as:
\begin{equation}
Z_i = C_{i+1} - C_i - spread;
\end{equation}
where the $spread = 0$, when opened long position is continued; \\
and $spread = cost$, when long position was opened after closing or opening a short position in case when there was no open position;\\
$cost$ is a variable that depends on the particular brokerage platform that was used for testing the strategy; in this case, for the currency pair EURUSD $cost = 0.00016$ was assumed and for the currency pair GBPUSD $cost = 0.00028$.
\item When short position is opened, profit after one candle (it does not mean that position is closed) can be calculated as: 
\begin{equation}
Z_i = C_i-C_{i+1}- spread;
\end{equation}
where the $spread = 0$, when opened short position is continued;\\ 
and $spread = cost$, when short position was opened after closing or opening a long position in case when there was no open position.
\end{itemize}

\begin{itemize}
\item If as the result of the test the number of red and green candles in the body of the earthworm had not obtained any dominance from the color that would recommend to open defined position (earthworm was white), it is assumed to continue the position, which is open, without any cost of the transaction (with the value of $spread = 0$). If at the beginning of the candle, at the time of checking the recommendation, there was no open position then it can be assumed to continue without any position.
\item Each of the profits calculated above is verified in terms of exceeding the allowable local position drawdown (within one candle). This additional risk management is widely used in automated trading and is called limiting losses --- Stop Loss ($SL$). In this algorithm, $SL$ was introduced as the fifth parameter of the strategy --- $p5$.
\end{itemize}

\noindent Presented strategy was tested if:
\begin{itemize}
\item in the case of opening at the beginning of the $i-th$ candle or continuation of a long position presented condition did not occur:\\
\begin{equation}
\text{If } L_i-O_i<SL \text{ then } Z_i=-SL
\end{equation}
where $L_i$ --– the lowest rate value in the $i-th$ candle; $SL$ --– absolute value Stop Loss appearing as $p5$ parameter;
\item in the case of opening at the beginning of the $i-th$ candle or continuation of a short position presented condition did not occur:
\begin{equation}
\text{If } O_i-H_i<SL \text{ then } Z_i=-SL
\end{equation}
where $H_i$ --– the highest rate value in the $i-th$ candle.
\end{itemize}
This way $p5$ parameter was used for securing in strategy the strong local changes in exchange rate within a single candle.\\
The meaning of $p5$ parameter to the behavior of the earthworm and the occurrence of the closing position is explained in Figure \ref{fig:fig5}.
\begin{figure}[h!]
\centering
\includegraphics[width = 0.5\textwidth]{figures/rys5.png}
\caption{Method of calculating the profit (here: loss) in the case of necessity to use $p5$ parameter actuating mechanism $SL$ for long positions.}
\label{fig:fig5}
\end{figure}
\FloatBarrier

\begin{itemize}
\item To determine the current quality of the strategy and risk management summation of current gains/losses (the accumulation) is performed
\begin{equation}
Zc_i = sum ( Z_i ) \text{ from } i = 1 \text{ to the current } i \text{ value;}
\end{equation}
\item There is a possible local subsidence $Zc_i$ capital by introducing two additional parameters to manage the risk.\\
$p3$ --- the number of steps back on the curve of the cumulative profit for the determination of $Zc_i$ --- $p3$ for comparison with the present value of $Zc_i$;
$p4$ --- maximum value measured in pips of the capital drawdown (eg for the currency pair EURUSD), which if exceeded will trigger the closure position:\\
\begin{equation}
\text{If } ( Zc_i - p3 - Z_i ) <- p4 \text{ and } PositionState \neq 0 \text{ then } PositionState = 0
\end{equation} 
where $PositionState$ --- is the type of open position (long or short) and $PositionState$ equal 0 means resignations from the opening position to the beginning of the next candle.
\end{itemize}
So, the above-described algorithm uses 5 parameters and 3 of them are devoted to risk management of the strategy.\\
%%PBStart
The complexity of the algorithm is illustrated in Figure \ref{fig:fig7}, which shows the individual steps and decisions made in the iteration on a single candle in the form of a decision tree. This iteration was carried out both in the search for optimal parameters $p1, ..., p5$ and in the process of testing the parameters on data that was not used in the learning period.
\begin{figure}[h!]
\centering
\includegraphics[width = \textwidth]{figures/rys7.png}
\caption{Decision tree for the states and decisions in one step (candle) in the learning and testing phase.}
\label{fig:fig7}
\end{figure}
\FloatBarrier

The figure shows in a tree form further decisions taken at the beginning of the candle and then reactions to any unacceptable change in course. The first step is to examine the state of the position which is taken from the closing of the previous one. The next step is to make decisions / recommendations as to the direction of the opening from the set \{long, short, none\}. After opening position  the return of the open position is observed --- it can be positive or negative. If it is positive, no any additional decisions are necessary. In the case of a negative return from the position both the level of $SL$ and a condition resulting from the use of parameters ($p3$, $p4$) are checked. The state at the end of tree is a beginning of the next candle.

Described algorithm was implemented in MATLAB environment and in Metatrader platform using the MQL4 language. The study was conducted both in the historical data (of course with the principles of machine learning --- testing was carried out using data not used in the learning phase) and a real ,,live'' market. Between the environments, there are important differences that affect the diversity of simulation results.

\section{Strategy studies}
The study was conducted primarily on the basic currency pair EURUSD at 1h candles. This market was chosen because of its popularity, a huge liquidity and also due to the fact that the authors posses many experiences in other machine learning strategies for this particular market such as in \cite{Wilinski2014,Wilinski}. These experiments suggested initially, carrying out many tests with a long learning period, about 1000 candles, then 500 candles and testing time windows of 50 to 200 candles. All these attempts have proven to be ineffective and led to continuing losses. After months of tests in many markets it was noted that over long periods of learning, e.g. the order of 1000 candles, variability in parameters was very large. Therefore, the tests started with a decidedly shorter periods of learning and testing.\\

Finally, the authors decided to present the results for the two markets --- currency pairs EURUSD and GBPUSD sampled at 1h --- in machine learning with ratios adjusted to the reduced scope of both the learning and testing phase.\\

>>>>

\mainmatter  % start of~an individual contribution

@@

% the name(s) of~the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and~the author index.
%
\author{Antoni Wiliński
\and Aneta Bera\and Maciej Jarlaczyński\and Piotr Błaszyński}
%
\authorrunning{A. Wilinski et al}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{West Pomeranian University of~Technology, Faculty of~Computer Science,\\
Zolnierska 49, 71210 Szczecin, Poland\\
\mailsa\\
\url{http://wi.zut.edu.pl}}

%
% NB: a more complex sample for affiliations and~the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%

\toctitle{The investment strategy based on behavior of~artificial earthworm for use in algotrading}
\tocauthor{Antoni Wilinski}
\maketitle

\begin{abstract}
The paper presents an innovative concept of~investment strategy derived from the~general ideas of~artificial intelligence. The strategy has been tested in a~number of~simulations on historical data as well as data outside learning periods. Tests were performed mainly in the~selected currency markets, including the~primary currency pair of~EURUSD. Presented approach opens long and~short positions alternately during candle closing events. Closing a~position is based on recommendation which is deduced from changes in characteristics of~time series. The strategy has parameters that vary through time and~are adapted to a~market changes. To present the~strategy a~metaphor of~an innovative artificial earthworm that feeds on a~time series was used. The earthworm has its own artificial intelligence which enables it to analyse ,,consumed'' candles, change its preferences about movement direction or~to stay in neutral state. The goal was to implement presented strategy for trading platform to achieve the~automatic trading effect.  The strategy was tested in two development environments --- MATLAB environment and~trading platform MetaTrader after conversion of~M-file into MQL4 file. Implemented strategy enables to achieve interesting results.
\keywords{investment strategy, machine learning, forecasting, time series, algotrading, mql4}
\end{abstract}

@@

Algotrading is a~new rapidly growing form of~trading on financial markets where investor (that makes decisions) is replaced with trading bot (computer program) \cite{Leshik2011,Wilinski2014}. Such machine can perform operations (opening and~closing positions, placing orders to open positions that meet certain conditions) directly in an automated brokerage platform environment. It may also occur as a~recommendation system that helps investor with making correct decisions.\\

Such computer applications (bots) require carefully prepared and~tested software that most often needs to adapt to the~changing situations on the~observed markets. Modern trends, in the~development of~this software, include sophisticated methods of: artificial intelligence, pattern recognition, classification and~machine learning \cite{Leshik2011,Wilinski,wang,sinclare}. Those methods are interpenetrating and~complementary and~are also expanded with information based on fundamental analysis e.g. automatically extracted from the~Internet \cite{elder,Wilinski2014,Schwager1996}. Many strategies are based on relatively simple rules. Efficiency of~these rules is amplified by~using of~machine learning \cite{Wilinski,person,tian,krutsinger,lewis,murphy}.\\

Strategy presented in this paper is an example of~a~solution that can be used fully automatically or~with assistance of~investor who makes the~decision manually. Using such softwaremakes sense only if there are reasonable premises of~achieving the~success. These conditions can be achieved through inductive thinking \cite{provost,KleskWilinski}, involving the~assumption that multiple validation of~decision on data from the~past substantiates the~correctness of~the~decision in the~future. If a~certain set of~rules in the~strategy used to seek the~pattern, leads to the~success then these rules are a~pattern that is encouraged for repeated use. If repeated use of~mentioned pattern results in more successes than failures, then the~basic premise to consider the~use of~the~pattern is rational \cite{bishop2006pattern,KleskWilinski}. Unfortunately, changing market is causing the~pattern to become outdated therefore it is necessary to constantly update it. Finding historically effective patterns that could be also effective in a~data space not used for their determination (to identify these patterns) is a~major challenge for machine learning which is applied in the~paper.
The article is organized as follows: at first authors describe an authored strategy based on an entirely innovative concept of~an artificial earthworm. Subsequently extensive studies of~effectiveness of~presented strategy in terms of~prediction and~the~aspect of~investment are carried out. At the~end the~conclusions resulting from the~study are presented and~also cited bibliographic sources are included.

\section{Strategy characteristics}
In presented strategy, a~metaphorical procedure is often used to explain heuristic algorithms of~artificial intelligence. The procedure is a~reference to an inspiration derived from the~nature.\\

Let's consider an artificial earthworm, which behaves like its natural inspirer. Let this artificial earthworm have a~length of~$p1$ candles (vectors representing the~knowledge about events occurring within the~contractual part of~the~time series). Let its movement come down to the~successive ,,eating'' one candle and~simultaneous ,,discarding'' of~one candle (the last one) --- always having $p1$ candles. Thus the~earthworm moves incrementally at length of~time called the~period of~the~candle. $i-th$ candle is  defined as a~row of~data of~values  $O_i$, $H_i$, $L_i$, $C_i$ (Open, High, Low and~Close). These values represent (in order): the~value at the~beginning of~this period, the~maximum value within the~period, the~minimum and~the~closing values - at the~end of~the~candle.\\
Let's assume that candle have a~color (often in a~number of~brokerage platforms such approach is used). Therefore candles may be red (when the~opening of~the~candle is higher than its closing $O_i>C_i$) and~green (when the~situation is opposite $O_i<C_i$).\\
Let's also assume --- it is a~primary strategy rule - that when the~artificial earthworm has a~significant dominance of~red candles, a~short position is opened or~continued. When earthworm has the~significant dominance of~green candles, a~long position is opened or~continued. Imprecise term ,,significant'' is defined by~the~use of~the~second parameter $p2$ according to the~rules:
\begin{equation}
\text{If }  Lr > p1/2 -p2 \text{ then } short;
\end{equation}
opening of~short position will be suggested;
\begin{equation}
\text{If }  Lg > p1/2 -p2 \text{ then } long;
\end{equation}
opening of~long position will be suggested;
\begin{equation}
\text{If }  Lg < p1/2 -p2 \text{ and~} Lr< p1/2-p2 \text{ then } none;
\label{eq:eq3}
\end{equation}
the continuation of~state will be suggested (position or~no opened position).\\

In the~rules above $Lr$ and~$Lg$ are the~number of~candles, respectively ,,red'' and~,,green'' in the~earthworm of~$p1$ length, wherein $p1 = Lr + Lg$.\\
When dominance determined by~another parameter $p2$ is undecided, as in (\ref{eq:eq3}), then earthworm takes the~color white and~the~position will not be opened. When the~earthworm continues to open position, investor does not incur costs. On many platforms, also in literature, symbolic marking of~growing candles in white and~decreasing in black is used. In this article, due to the~occurrence of~three values, green-red-white colors are used.\\

The earthworm is therefore supplied with artificial intellect which enables the~accumulation of~knowledge and~decision-making:
\begin{itemize}
\item right after closing of~the~current candle, earthworm ,,swallows'' the~candle and~,,knows'' what color is it;
\item earthworm knows how many candles it is composed with;
\item after absorbing and~discarding one candle, earthworm knows how many candles inside it are red or~green and~how to determine its status --- red, green or~white;
\item knowing its status, earthworm knows how to make a~decision about opening or~continuing position; the~opening of~a~position (long or~short) means the~simultaneous closing of~the~current position (opposite) if such position is opened;
\item earthworm observes cumulative profit curve and~if it  notices unacceptable drawdown, then opened position is changed to the~opposite;
\item unacceptable drawdown is defined by~two other parameters $p3$ and~$p4$. $p3$ is the~number of~previous candles, which sets a~reference point in time series, in which the~value of~the~accumulated profit for $i-p3$ is checked for comparison with the~current value of~the~accumulated profit. $p4$ is permissible decrease of~cumulative profit. If it is exceeded, the~$SL$ (Stop Loss --- closes the~currently open position) mechanism is activated;
\item earthworm also observes current course changes during  each next candle and~reacts to unprofitable changes. If the~current exchange rate causes a~loss on the~opened positions greater than $p5$ then the~$SL$ mechanism is triggered --- the~position is closed and~till the~end of~the~candle earthworm remains in the~state without an open position. $p5$ is therefore the~fifth strategy parameter. 
\end{itemize}

These features of~the~earthworm's intelligence may be extended, however, for this work they were limited to the~foregoing. \\

It is assumed that the~time series is $S_i$ where $i = 1, 2, ...,$ and~each element is a~vector $S_i = (O_i, H_i, L_i, C_i)$ called candle.\\

Artificial earthworm executes the~following algorithm.\\
The strategy is built on the~principle of~classical machine learning. Before starting to run the~algorithm on data located outside the~scope that investor knows (like on the~current market segments --- out-of-sample), it is necessary to ,,teach''  the~values of~parameters. This principle is presented in Figure \ref{fig:fig2}, which shows examples of~learning periods and~following them testing periods. Data from testing period was not used during learning.\\
\begin{figure}[ht]
\centering
\includegraphics[width = 0.6\textwidth]{figures/rys2.png}
\caption{An example of~the~arrangement of~learning and~testing periods of~strategy parameters in time series.}
\label{fig:fig2}
\end{figure}
\FloatBarrier
\vspace{-1em}
This repeatable algorithm located in both periods - learning and~testing - has the~following structure. In learning period the~optimal value for five parameters $p1$, $p2$, ..., $p5$ is determined. In test period these designated values of~parameters are used to perform the~algorithm in the~new environment. In each $i-th$ step (candle) of~time series following activities are performed:
\begin{itemize}
\item For each $i = p1, p1 +1, p1 +2, ..., I$ and~for $j = 1, ..., p1$ check whether the~candle located at a~distance $i-j$ backwards from the~current candle is:\\
green when $O_{i-j} <C_{i-j}$, \\
red when $O_{i-j}> C_{i-j}$, \\
where $p1$ --- is the~first parameter which is the~length of~the~artificial earthworm --- the~number of~candles of~different colors that make up the~,,body'' of~the~earthworm; \\
$O_{i-j}$ --- is the~opening value of~(i-j)-th candle;\\ 
$C_{i-j}$ --- is the~closing value of~(i-j)-th candle; \\
$I$ --- is the~length of~the~test window, after which it is necessary to change the~parameters values to the~new, learned in the~machine learning phase;\\
\item In each $i-th$ step, the~loop $j = 1, ..., p1$ is performed in which a~number of~green $Lg$, red $Lr$ candles is calculated so the~sum of~$Lg$ and~$Lr$ is $p1$ (to avoid the~situation when $O_i = C_i$ is an arbitrary assignation to one of~the~candles sets e.g. green);
\item A condition for the~existence of~a~$Rg$ recommendation to open a~long position (or to continue it if it is open) is checked. This condition is arbitrarily adopted (one can of~course consider different one): 
\begin{equation}
\text{If } Lg> p1/2-p2 \text{ then } Rg=1
\end{equation}

\end{itemize}
If this condition is fulfilled, then depending on the~current position status $PositionState$ (where $PositionState$ is a~variable which takes values from \{$long, short, none$\}) at the~opening of~the~next candle, a~position is opened or~continued as follows: 
\begin{equation}
\text{If } Rg=1 \text{ and~} PositionState =long \text{ then }PositionState =long;
\end{equation}
(costless continuation of~opened position);
\begin{equation}
\text{If } Rg=1 \text{ and~} PositionState = short \text{ then } PositionState=long;
\end{equation}
\begin{equation}
\text{If } Rg=1 \text{ and~} PositionState =none \text{ then } PositionState =long;
\end{equation}

Similarly check the~condition of~recommendations Rr to open or~continue short position:
\begin{equation}
\text{If } Lr>p1/2 - p2 \text { then Rr=1.}
\end{equation}
If this condition is fulfilled, then depending on what the~current position status PositionState is, at the~opening of~the~next candle, position is opened or~continued as follows: \\
\begin{equation}
\text{If } Rr=1 \text{ and~} PositionState =long \text{ then } PositionState =short;
\end{equation}
\begin{equation}
\text{If } Rr==1 \text{ and~} PositionState ==short \text{ then } PositionState=short;
\end{equation}
(costless continuing of~opened position)
\begin{equation}
\text{If } Rr==1 \text{ and~}  PositionState ==none \text{ then }  PositionState =short;
\end{equation}

In formulas above there is a~parameter $p2$. It is the~second parameter of~the~strategy which is an arbitrarily separated part of~the~earthworm, which depending on size causes the~direction of~investment not to change. Large $p2$ causes long positions to change into short positions (or vice-versa) rarely. Small value of~$p2$ leads to a~higher frequency of~these changes. The optimum value of~$p2$ is chosen similarly like $p1$.
\begin{itemize}
\item 	When opening long position, profit after one candle (this does not mean that position was closed after a~single candle) can be calculated as:
\begin{equation}
Z_i = C_{i+1} - C_i - spread;
\end{equation}
where the~$spread = 0$, when opened long position is continued; \\
and $spread = cost$, when long position was opened after closing or~opening a~short position in case when there was no open position;\\
$cost$ is a~variable that depends on the~particular brokerage platform that was used for testing the~strategy; in this case, for the~currency pair EURUSD $cost = 0.00016$ was assumed and~for the~currency pair GBPUSD $cost = 0.00028$.
\item When short position is opened, profit after one candle (it does not mean that position is closed) can be calculated as: 
\begin{equation}
Z_i = C_i-C_{i+1}- spread;
\end{equation}
where the~$spread = 0$, when opened short position is continued;\\ 
and $spread = cost$, when short position was opened after closing or~opening a~long position in case when there was no open position.
\end{itemize}

\begin{itemize}
\item If as the~result of~the~test the~number of~red and~green candles in the~body of~the~earthworm had not obtained any dominance from the~color that would recommend to open defined position (earthworm was white), it is assumed to continue the~position, which is open, without any cost of~the~transaction (with the~value of~$spread = 0$). If at the~beginning of~the~candle, at the~time of~checking the~recommendation, there was no open position then it can be assumed to continue without any position.
\item Each of~the~profits calculated above is verified in terms of~exceeding the~allowable local position drawdown (within one candle). This additional risk management is widely used in automated trading and~is called limiting losses --- Stop Loss ($SL$). In this algorithm, $SL$ was introduced as the~fifth parameter of~the~strategy --- $p5$.
\end{itemize}

\noindent Presented strategy was tested if:
\begin{itemize}
\item in the~case of~opening at the~beginning of~the~$i-th$ candle or~continuation of~a~long position presented condition did not occur:\\
\begin{equation}
\text{If } L_i-O_i<SL \text{ then } Z_i=-SL
\end{equation}
where $L_i$ --– the~lowest rate value in the~$i-th$ candle; $SL$ --– absolute value Stop Loss appearing as $p5$ parameter;
\item in the~case of~opening at the~beginning of~the~$i-th$ candle or~continuation of~a~short position presented condition did not occur:
\begin{equation}
\text{If } O_i-H_i<SL \text{ then } Z_i=-SL
\end{equation}
where $H_i$ --– the~highest rate value in the~$i-th$ candle.
\end{itemize}
This way $p5$ parameter was used for securing in strategy the~strong local changes in exchange rate within a~single candle.\\
The meaning of~$p5$ parameter to the~behavior of~the~earthworm and~the~occurrence of~the~closing position is explained in Figure \ref{fig:fig5}.
\begin{figure}[h!]
\centering
\includegraphics[width = 0.5\textwidth]{figures/rys5.png}
\caption{Method of~calculating the~profit (here: loss) in the~case of~necessity to use $p5$ parameter actuating mechanism $SL$ for long positions.}
\label{fig:fig5}
\end{figure}
\FloatBarrier

\begin{itemize}
\item To determine the~current quality of~the~strategy and~risk management summation of~current gains/losses (the accumulation) is performed
\begin{equation}
Zc_i = sum ( Z_i ) \text{ from } i = 1 \text{ to the~current } i \text{ value;}
\end{equation}
\item There is a~possible local subsidence $Zc_i$ capital by~introducing two additional parameters to manage the~risk.\\
$p3$ --- the~number of~steps back on the~curve of~the~cumulative profit for the~determination of~$Zc_i$ --- $p3$ for comparison with the~present value of~$Zc_i$;
$p4$ --- maximum value measured in pips of~the~capital drawdown (eg for the~currency pair EURUSD), which if exceeded will trigger the~closure position:\\
\begin{equation}
\text{If } ( Zc_i - p3 - Z_i ) <- p4 \text{ and~} PositionState \neq 0 \text{ then } PositionState = 0
\end{equation} 
where $PositionState$ --- is the~type of~open position (long or~short) and~$PositionState$ equal 0 means resignations from the~opening position to the~beginning of~the~next candle.
\end{itemize}
So, the~above-described algorithm uses 5 parameters and~3 of~them are devoted to risk management of~the~strategy.\\
%%PBStart
The complexity of~the~algorithm is illustrated in Figure \ref{fig:fig7}, which shows the~individual steps and~decisions made in the~iteration on a~single candle in the~form of~a~decision tree. This iteration was carried out both in the~search for optimal parameters $p1, ..., p5$ and~in the~process of~testing the~parameters on data that was not used in the~learning period.
\begin{figure}[h!]
\centering
\includegraphics[width = \textwidth]{figures/rys7.png}
\caption{Decision tree for the~states and~decisions in one step (candle) in the~learning and~testing phase.}
\label{fig:fig7}
\end{figure}
\FloatBarrier

The figure shows in a~tree form further decisions taken at the~beginning of~the~candle and~then reactions to any unacceptable change in course. The first step is to examine the~state of~the~position which is taken from the~closing of~the~previous one. The next step is to make decisions / recommendations as to the~direction of~the~opening from the~set \{long, short, none\}. After opening position  the~return of~the~open position is observed --- it can be positive or~negative. If it is positive, no any additional decisions are necessary. In the~case of~a~negative return from the~position both the~level of~$SL$ and~a~condition resulting from the~use of~parameters ($p3$, $p4$) are checked. The state at the~end of~tree is a~beginning of~the~next candle.
